Inferring social kernels and segregation measures from ego-network data

### 2020-05-12 main.tex
Prior to Nick's comments.

\documentclass{scrartcl}
\usepackage{authblk}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amsthm}
\usepackage{csquotes}
\usepackage[british]{babel}
\usepackage[giveninits,sorting=none,style=numeric-comp]{biblatex}
\usepackage{hyperref}
\usepackage{tabularx}
\usepackage{booktabs}
\usepackage{cleveref}
\usepackage{fmtcount}
\usepackage{graphicx}
\DeclareGraphicsExtensions{.pdf}

\newcommand{\expit}{\sigma}
\newcommand{\transpose}[1]{#1^\intercal}

@@

  We fuse the notion of Blau space---a space spanned by the attributes of the members of society that inhabit it---with conditionally-independent edge models to obtain a generative model for social networks. Under a Bayesian paradigm, we develop the techniques to infer the parameters of the model from ego network data. The model-based approach allows us to develop a social segregation measure with desirable properties, including an intuitive scale, being applicable to multivariate and mixed node attributes, as well as being defined at the level of individuals and the society as a whole. For pairs of individuals, the segregation measure induces a semi-metric that can be used to measure the social distance between them in a principled fashion.

  We apply our approach to ego network data collected in the United Kingdom and United States from four surveys, and make recommendations for future ego-network surveys. The  importance of different dimensions of Blau space is similar across time and location, suggesting a macroscopically stable social fabric. Physical separation and age differences have the most significant impact on social distance within friendship networks with implications for intergenerational mixing and isolation in later stages of life.
\end{abstract}

\section{Introduction}

Homophily, the tendency for people to connect with others who are alike, is one of the most robust observations of the social sciences and shapes how our society is connected~\cite{McPherson2001}. Whilst institutionalised segregation and discrimination have largely been outlawed in the developed world, voluntary association between individuals who share common traits makes social circles more homogeneous than one would expect by chance. Quantifying homophily is not only important for understanding why social ties form between some people yet not between others, but the manifestation of homophily as poorly-connected social networks can have a significant impact on dynamics unfolding on the social fabric~\cite{Golub2012}. For example, users of online social networks such as Facebook and Twitter tend to connect with others who hold similar political views~\cite{Boutyline2017}. They are more likely to be exposed to information that confirms rather than challenges their beliefs~\cite{Bakshy2015}. An ``echo chamber'' effect ensues, leading to polarised opinions~\cite{DeMarzo2003}. Homophily can also have a detrimental impact on public health: \textcite{Salathe2008} showed that clusters of individuals who mutually reinforce their belief that vaccinations are harmful can raise the likelihood of significant disease outbreaks---even if the vaccination rate is above herd immunity levels on average.

Homophily can be observed in friendships~\cite{Currarini2009, Hipp2009}, networks of discussion partners~\cite{McPherson2006}, communication networks~\cite{Wang2013, Leo2016}, marital ties~\cite{Blau1984}, and online social networks~\cite{Chang2010}. Relationships are homogeneous with respect to a wide range of attributes such as age~\cite{Marsden1988,Smith2014}, gender~\cite{Stehle2011,Smith2014}, ethnicity~\cite{Chang2010, Blumenstock2013, Currarini2009}, education~\cite{McPherson2006, Smith2014,Johnson1989}, occupation~\cite{Chan2004}, income~\cite{Leo2016, Wang2013, Johnson1989}, religion~\cite{Platt2012}, parental~\cite{Johnson1989} and marital status~\cite{Kalmijn2007}, political ideology~\cite{Bakshy2015, Boutyline2017}, and geographical location~\cite{Lambiotte2008, Expert2011, Backstrom2010, Scellato2011, Illenberger2013}. Romantic relationships, which tend to be heterophilous with respect to gender, are a notable exception.

The study of homophily is complicated by the scarcity of high-quality data~\cite{Butts2012, Blumenstock2013}: we need social network data together with demographic information for each person. On the one hand, online social networks and the widespread use of mobile phones provide us with detailed information about connections between individuals~\cite{Golder2014}, and seemingly private traits such as socioeconomic status~\cite{Blumenstock2015, Luo2017}, sexual orientation~\cite{Wang2017}, age, gender, and political ideology can be inferred~\cite{Kosinski2013}. Unfortunately, network features are often used to predict demographic attributes~\cite{Wang2013,Blumenstock2015,Luo2017,Kosinski2013}, which would confound any study of homophily: we cannot use the same data to predict demographic attributes and study homophily with respect to those attributes. Furthermore, data are too revealing in terms of privacy but, at the same time, do not provide enough information for researchers~\cite{Golder2014}. Individuals can be identified in anonymised social networks~\cite{Backstrom2011, Narayanan2008}, and augmenting the network data with demographic information would make re-identification even easier.

@@

% In this paper, we aim to address three challenges. First, numerous studies~\cite{Cheadle2012, Currarini2009, Hipp2009,Marsden1987, Marsden1988, McPherson2001, Smith2014, Wimmer2010} refer to Blau's~\cite{Blau1977} intuitive and powerful theory of social structure: people occupy a multidimensional social space spanned by their demographic attributes, and their rates of interaction depend on their relative positions. However, to the best of our knowledge, only \textcite{Hipp2009} explicitly consider such a social space---also referred to as ``Blau space''~\cite{McPherson2001}. \tah{need to double check that Hipp is indeed the only one; Hoff's 2002 paper also had something like that in it}

@@

Building on the successes of latent space models for social networks~\cite{Hoff2002,Handcock2007,Raftery2012}, we consider a generative model for social networks whose members occupy a multidimensional Blau space. We derive the posterior for model parameters given partial observations of such social networks obtained from surveys and show that the posterior is equivalent to the widely-known ``prior correction'' when the probability for two people to connect is modelled by a logistic model. We apply our approach to nine existing datasets from the United Kingdom (UK) and two from the United States (US). Our analysis reveals that the effects of homophily on society are remarkably stable in both countries regardless of time and the specific nature of relationships. We provide recommendations for conducting surveys to infer homophily in social networks and discuss future work, including principled imputation of demographic attributes in social networks and using the generative model to generate synthetic networks: an approach that has attracted interest to circumvent the difficulties associated with anonymising social network data~\cite{Pfeiffer-III2014, Lieberman2010, Lieberman2012, Nettleton2016, Sathanur2017}.

We also discuss desirable properties for social segregation measures and, using the generative network model, develop a suite of measures applicable to arbitrary attributes. The measures capture segregation at different scales: pairs of individuals, single individuals, and society as a whole. We illustrate the measures with a simple example, and we show that it reduces to a well-known segregation measure if the attributes are univariate and categorical. We apply the segregation measures to survey data, finding that physical separation and age are the most important factors contributing to the segregation of society.

@@

\subsection{Generative network model and observation\label{sec:model}}

\begin{figure}
    \includegraphics{survey-synthetic}
    \titlecaption{A simple generative model for ego networks obtained from surveys.}{Panel~(a) shows synthetic survey data generated by our model. Seeds $\seeds$ are represented by blue squares, nominees $\nominees$ by orange circles, and the remainder of the population $\missing$ by grey circles. Panel~(b) illustrates the parts of the adjacency matrix revealed by a particular two-stage survey. Connections are represented by black entries in the adjacency matrix, absent connections by white entries, and lacking information by grey entries.\label{fig:survey-synthetic}}
\end{figure}

We build a generative model for ego networks in two steps: we discuss a model for entire social networks first and the data collection process second. In particular, we consider a population of $n$ individuals $\population$ who occupy a Blau space $\blauspace$ spanned by their demographic attributes, such as age, income, or gender. In contrast to common latent space models~\cite{Hoff2002, Hoff2008, Handcock2007}, the attributes are observed although \textcite{Hoff2002} also consider observed covariates. This allows us to learn how members of the population connect with one another. In particular, the $q$-dimensional attribute vector $x_i\in\blauspace$ for each individual $i\in \population$ is drawn independently from a distribution of demographic attributes $P(x_i)$. Connections between individuals are encoded by the binary adjacency matrix $A$ such that $A_{ij}=1$ if $j$ considers $i$ to be a friend and $A_{ij}=0$ otherwise. We assume that people do not interact with themselves such that $A_{ii}=0$ for all $i$, and that connections are directed because social ties need not be reciprocated. For example, Alice may consider Bob to be a friend whereas the reverse need not be true~\cite{Ball2013}.

Given the positions of two individuals $i$ and $j$ in Blau space, we assume that connections form independently with probability $\rho(x_i, x_j,\theta)$, i.e.\ we consider a spatial network model~\cite{Barnett2007} with conditionally-independent edges~\cite{Fienberg2012}. The \emph{connectivity kernel} $\rho$ is parametrised by a vector of parameters $\theta$ which we would like to infer. The assumption of conditionally-independent edges can be problematic. For example, it is difficult to reproduce some of the statistics of real-world networks, such as heavy-tailed degree distributions, if the node density is homogeneous~\cite{Barnett2007}. Furthermore, keeping $\theta$ constant, the average degree scales linearly with the number of nodes~\cite{Caron2017}. Nevertheless, we use spatial network models because the connectivity kernel is easily interpretable and they are able to capture salient features of social networks. For example, nodes in high-density regions have larger degrees on average~\cite{Barnett2007}. Similarly, members of the ethnic majority have more social ties in social networks in US high schools~\cite{Currarini2009}.

We now turn to the data collection process. Observations are collected in two stages: first, we sample a subset $\seeds\subset\population$ of \emph{seeds} uniformly at random and collect information about their demographic background. We also ask each seed $j\in \seeds$ about their social ties $\nominees_j=\{i\in\population:A_{ij}=1\}$, henceforth \emph{nominees}. Second, we collect demographic information about the nominees $\nominees=\bigcup_{j\in \seeds} \nominees_j$ by asking the seeds about their friends~\cite{Huckfeldt1983, McPherson2006, Kalmijn2007} or conducting follow-up surveys~\cite{Johnson1989}.

Panel~(a) of \cref{fig:survey-synthetic} illustrates the model with a simple example: we first draw the positions $x$ of $n=1,000$ nodes in the social space uniformly at random from the two-dimensional unit box. Second, we connect nodes with each another using a connectivity kernel that induces homophily along the two dimensions of Blau space. Finally, we sample $\card{\seeds}=100$ seeds from the population, and collect information about their demographic background as well as their $\card{\nominees}=108$ nominees, which allows us to embed the ego networks in Blau space.

\subsection{Inference}

\tah{big gaping hole here}

\subsection{Developing a segregation measure}

The study of homophily is intimately tied to measuring segregation, and a wide range of measures have been developed (see \textcite{Rodriguez-Moral2016, Bojanowski2014} for reviews). Many approaches are based on co-presence in organisational units such as schools~\cite{Orfield2014}, voluntary associations~\cite{Popielarz1999}, occupations~\cite{Charles1995}, or census tracts~\cite{Reardon2004,Reardon2011}, and we refer to them as \emph{organisational} measures. Typically, they compare how the distribution of demographic attributes within organisational units differs from the distribution of attributes in the general population. Whilst organisational measures are applicable to any data that can be stratified according to a variable of interest, they cannot capture segregation at smaller scales than the strata~\cite{Blumenstock2013}. For example, the ethnic composition of a set of schools may be representative of the general population indicating that there is no (organisational) segregation. But the social networks \emph{within} the schools may exhibit strong ethnic homophily~\cite{Currarini2009,Moody2001}.
%Similarly, social networks exhibit income homophily~\cite{Leo2016}, but neighbourhoods are often heterogeneous with respect to income~\cite{Ioannides2004}.
Organisational measures cannot capture such social segregation.

\emph{Social} measures of segregation, such as the assortativity coefficient~\cite{Newman2003a}, overcome these limitations by explicitly considering the interactions amongst individuals~\cite{Blumenstock2013}, but have their own difficulties: first, they usually rely on the existence of mutually exclusive groups such as sex, ethnicity, or religion~\cite{Bojanowski2014}, and they are not applicable to continuous attributes such as age or income. Attributes are often discretised~\cite{Lam-Morgan2012, Kalmijn2007, Kim2012},
%and models for ``ordered community structure'' in networks have been developed~\cite{Gregory2012},
but the boundaries between categories always suffer from some degree of arbitrariness~\cite{Reardon2004}. Second, segregation for multiple attributes can be measured independently, but it is unclear how to define a composite segregation measure. Furthermore, social measures are typically defined as summary statistics of a fully-observed social network. Consequently, we cannot easily quantify uncertainties, and it is non-trivial to apply them to sampled network data.

In addition to addressing the challenges above, a desirable social segregation measure should satisfy the following properties: first, the measure should be insensitive to the overall edge density to facilitate comparison of segregation across different networks. Otherwise, the segregation measure would depend on the size of the population because the edge density scales as $n^{-1}$ if the average degree is approximately constant. Second, following \textcite{Freeman1978}, we would like the measure to capture the notion that segregation places ``restrictions on the access of people to one another''. Third, the measure should be easily interpretable, and it should have a natural notion of the absence of segregation when individuals form connections without regard to their positions in Blau space. For example, the difference of within- and between-group ties considered by \textcite{Krackhardt1988} depends on the sizes of the groups even if there is no homophily: there is no natural reference point.

A single measure cannot capture the complexities of social networks, and we will develop a suite of measures applicable at different scales: the segregation or separation between any two individuals, the isolation experienced by any one individual, and a global measure of segregation. Starting at the microscopic level, we use the generative model discussed in \cref{sec:model} to define a \emph{pairwise} measure of separation between two individuals with attributes $x$ and $y$
\begin{equation}
    \varphi(x, y) = -\log\frac{\rho(x, y)}{\rho(y, y)},\label{eq:segregation-pairwise}
\end{equation}
where $\rho(x, y)$ is the probability for $y$ to connect with $x$, and, for notational convenience, we use $x$ to refer to an individual with attributes $x$ as well as the attributes themselves. The probability $\rho(y, y)$ for $y$ to connect with someone with identical attributes serves as a reference point such that the measure does not depend on the overall edge density. An individual $y$ is $\exp(\varphi(x, y))$ times less likely to connect with someone with attributes $x$ than they are to connect with someone with identical attributes $y$, and $\varphi$ may be understood as the social isolation experienced by $x$ as a result of the behaviour of $y$. The measure is zero if people do not discriminate with respect to the attributes or if two individuals have the same demographic attributes. In a heterophilous society, the measure is negative for any pair of individuals. For a homophilous connectivity kernel, the measure is positive and is a \emph{semi-metric} for Blau space; we will consider a family of connectivity kernels for which $\varphi$ is a true metric in \cref{sec:segregation-distance}.

%\[
%    \varphi(x, y) = -\log\frac{\rho(x, y)(1-\rho(y, y))}{\rho(y, y)(1-\rho(x,y))}
%\]

\begin{prop}\label{prop:segregations-semi-metric}
    If the connectivity kernel is homophilous, symmetric, and homogeneous, the pairwise segregation measure is a \emph{semi-metric}~\cite{Wilson1931}, i.e.\ it satisfies the properties of a metric except the triangle inequality.
\end{prop}
\begin{proof}
    First, the measure is zero for any two individuals with the same attributes by substitution into \cref{eq:segregation-pairwise}. Second, the measure is non-negative because homophily implies that $\rho(x, y)<\rho(y, y)$. Third, the numerator of \cref{eq:segregation-pairwise} is invariant under exchange of the arguments, and the denominator is constant because the kernel is symmetric and homogeneous, respectively. The triangle inequality is not necessarily satisfied.
\end{proof}

The less likely two people are to connect, the larger the social distance between them. The assumptions required for \cref{prop:segregations-semi-metric} to hold may seem restrictive, but they are satisfied by most studies of spatial networks~\cite{Butts2012,Lambiotte2008,Expert2011}.

@@

For example, consider a stochastic block model (SBM)~\cite{Snijders2011} with intra-group connection probability $\rho_\mathrm{same}$ and inter-group connection probability $\rho_\mathrm{different}<\rho_\mathrm{same}$. The pairwise segregation measure
\[
    \varphi(x, y) = -\left(1-\delta_{xy}\right)\log\frac{\rho_\mathrm{different}}{\rho_\mathrm{same}},
\]
where $x$ and $y$ denote the blocks and $\delta_{xy}$ is the Kronecker delta. The segregation measure only depends block membership, and it does not depend on the size of each block. The pairwise measure for members of different block is the log odds ratio for the existence of intra-group ties as opposed to inter-group ties, which has been considered previously by \textcite{Moody2001}.

Both the connectivity kernel and the pairwise measure convey information about how two individuals connect with one another. To quantify segregation at the level of an individual $x$, we define the \emph{individual} segregation measure or \emph{social isolation}
\begin{equation}
    \phi(x) = \int dy \  P(y) \varphi(x, y),\label{eq:segregation-individual}
\end{equation}
which measures the average distance from members of society to $x$. For our SBM,
\begin{equation}
    \phi(x) = -(1-P(x))\log\frac{\rho_\mathrm{different}}{\rho_\mathrm{same}},\label{eq:segregation-individual-alien}
\end{equation}
and members of all blocks experience the same degree of isolation if the blocks are of the same size. If the sizes are unequal, minorities experience more isolation and majority groups experience less isolation. Indeed, ethnic minorities in schools tend to be more isolated and have fewer social ties~\cite{Currarini2009}.

@@

Some of the demographic attributes of seeds and nominees are missing because respondents did not know or refused to provide the information, and we drop individuals with one or more missing attributes, as shown in \cref{tbl:survey-sample-size}. Such a complete-case analysis can introduce biases if the data are not missing completely at random~\cite{Rubin1976}, but handling the missing data in a principled fashion would, once again, require us to develop a model for demographic attributes~\cite{Pigott2001, Gelman2013}.

>>>>

\documentclass{scrartcl}
\usepackage{authblk}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amsthm}
\usepackage{csquotes}
\usepackage[british]{babel}
\usepackage[
    giveninits,
    backend=biber,
    sorting=none,
    citestyle=numeric-comp,
    maxnames=3,
]{biblatex}
\usepackage{hyperref}
\usepackage{tabularx}
\usepackage{booktabs}
\usepackage{cleveref}
\usepackage{fmtcount}
\usepackage{graphicx}
\DeclareGraphicsExtensions{.pdf}

\usepackage[usenames]{xcolor}
\definecolor{mpl0}{HTML}{1f77b4}
\definecolor{mpl1}{HTML}{ff7f0e}
\definecolor{mpl2}{HTML}{2ca02c}
\definecolor{mpl3}{HTML}{d62728}
\definecolor{mpl4}{HTML}{9467bd}
\definecolor{mpl5}{HTML}{8c564b}
\definecolor{mpl6}{HTML}{e377c2}
\definecolor{mpl7}{HTML}{7f7f7f}
\definecolor{mpl8}{HTML}{bcbd22}
\definecolor{mpl9}{HTML}{17becf}

\hypersetup{
    colorlinks=true,
    linkcolor=mpl0,
    citecolor=mpl0,
    urlcolor=mpl0,
}

\newcommand{\expit}{\sigma}
\newcommand{\transpose}[1]{#1^\intercal}
\newcommand{\observed}[1]{#1}
\DeclareMathOperator{\logit}{logit}
\DeclareMathOperator{\logp}{log1p}
\DeclareMathOperator{\logsumexp}{logsumexp}

@@

  We fuse the notion of Blau space---a space spanned by the attributes of the members of society that inhabit it---with conditionally-independent edge models to obtain a generative model for social networks. The model-based approach allows us to develop a social segregation measure with desirable properties, including an intuitive scale, being applicable to multivariate and mixed node attributes, as well as being defined at the level of individuals and the society as a whole. For pairs of individuals, the segregation measure induces a semi-metric that can be used to measure the social distance between them in a principled fashion.

  Under a Bayesian paradigm, we develop the techniques to infer the parameters of the model from ego network data. We apply our approach to eleven ego network datasets collected in four surveys in the United Kingdom and United States, and we make recommendations for future ego-network surveys. The importance of different dimensions of Blau space is similar across time and location, suggesting a macroscopically stable social fabric. Physical separation and age differences have the most significant impact on social distance within friendship networks with implications for intergenerational mixing and isolation in later stages of life.
\end{abstract}

\section{Introduction\label{sec:introduction}}

Homophily, the tendency for people to connect with others who are alike, is one of the most robust observations of the social sciences and shapes how our society is connected~\cite{McPherson2001}.
%Whilst institutionalised segregation and discrimination have largely been outlawed in the developed world, voluntary association between individuals who share common traits makes social circles more homogeneous than one would expect by chance.
Quantifying homophily is not only important for understanding why social ties form between some people yet not between others, but the manifestation of homophily as poorly-connected social networks can have a significant impact on dynamics unfolding on the social fabric~\cite{Golub2012}. For example, users of online social networks such as Facebook and Twitter tend to connect with others who hold similar political views~\cite{Boutyline2017}. They are more likely to be exposed to information that confirms rather than challenges their beliefs~\cite{Bakshy2015}. An ``echo chamber'' effect ensues, leading to polarised opinions~\cite{DeMarzo2003}. Homophily can also have a detrimental impact on public health: \textcite{Salathe2008} showed that clusters of individuals who mutually reinforce their belief that vaccinations are harmful can raise the likelihood of significant disease outbreaks---even if the vaccination rate is above herd immunity levels on average.

Homophily can be observed in friendships~\cite{Currarini2009, Hipp2009}, networks of discussion partners~\cite{McPherson2006}, communication networks~\cite{Wang2013, Leo2016}, marital ties~\cite{Blau1984}, and online social networks~\cite{Chang2010}. Relationships are homogeneous with respect to a wide range of attributes, including age~\cite{Marsden1988,Smith2014}, gender~\cite{Stehle2011,Smith2014}, ethnicity~\cite{Chang2010, Blumenstock2013, Currarini2009}, education~\cite{McPherson2006, Smith2014,Johnson1989}, occupation~\cite{Chan2004}, income~\cite{Leo2016, Wang2013, Johnson1989}, religion~\cite{Platt2012}, parental~\cite{Johnson1989} and marital status~\cite{Kalmijn2007}, political ideology~\cite{Bakshy2015, Boutyline2017}, and geographical location~\cite{Lambiotte2008, Expert2011, Backstrom2010, Scellato2011, Illenberger2013}. % Romantic relationships, which tend to be heterophilous with respect to gender, are a notable exception.

Segregation measures are often used to quantify homophily~\cite{Rodriguez-Moral2016, Bojanowski2014}. Many approaches are based on co-presence in organisational units such as schools~\cite{Orfield2014}, voluntary associations~\cite{Popielarz1999}, occupations~\cite{Charles1995}, or census tracts~\cite{Reardon2004,Reardon2011}, and we refer to them as \emph{organisational} measures. Typically, they compare how the distribution of demographic attributes within organisational units differs from the distribution of attributes in the general population. Whilst organisational measures are applicable to any data that can be stratified according to a variable of interest, they cannot capture segregation at smaller scales than the strata~\cite{Blumenstock2013}. For example, the ethnic composition of a set of schools may be representative of the general population indicating that there is no (organisational) segregation. But the social networks \emph{within} the schools may exhibit strong ethnic homophily~\cite{Currarini2009,Moody2001}.
%Similarly, social networks exhibit income homophily~\cite{Leo2016}, but neighbourhoods are often heterogeneous with respect to income~\cite{Ioannides2004}.
Organisational measures cannot capture such social segregation.

\emph{Social} measures of segregation, such as the assortativity coefficient~\cite{Newman2003a}, overcome these limitations by explicitly considering the interactions amongst individuals~\cite{Blumenstock2013}, but have their own difficulties: first, they usually rely on the existence of mutually exclusive groups such as sex, ethnicity, or religion~\cite{Bojanowski2014}, and they are not applicable to continuous attributes such as age or income. Attributes are often discretised~\cite{Lam-Morgan2012, Kalmijn2007, Kim2012},
%and models for ``ordered community structure'' in networks have been developed~\cite{Gregory2012},
but the boundaries between categories always suffer from some degree of arbitrariness~\cite{Reardon2004}. Second, segregation for multiple attributes can be measured independently, but it is unclear how to define a composite segregation measure. Furthermore, social measures are typically defined as summary statistics of a fully-observed social network. Consequently, we cannot easily quantify uncertainties. %, and it is non-trivial to apply them to sampled network data.

Building on the successes of conditionally-independent edge models~\cite{Snijders2011} and, in particular, latent space models for social networks~\cite{Hoff2002,Handcock2007,Raftery2012}, we consider a generative model for social networks whose members occupy a multidimensional Blau space. We discuss desirable properties for social segregation measures, and, using the generative network model, we develop a suite of measures applicable to arbitrary attributes. The measures capture segregation at different scales: pairs of individuals, single individuals, and society as a whole. We illustrate the measures with a simple example, and we show that it reduces to a well-known segregation measure if the attributes are univariate and categorical.

In practice, the study of homophily is complicated by the scarcity of high-quality data~\cite{Butts2012, Blumenstock2013}: we need social network data together with demographic information for each person. On the one hand, online social networks and the widespread use of mobile phones provide us with detailed information about connections between individuals~\cite{Golder2014}, and seemingly private traits such as socioeconomic status~\cite{Blumenstock2015, Luo2017}, sexual orientation~\cite{Wang2017}, age, gender, and political ideology can be inferred~\cite{Kosinski2013}. Unfortunately, network features are often used to predict demographic attributes~\cite{Wang2013,Blumenstock2015,Luo2017,Kosinski2013}, which would confound any study of homophily: we cannot use the same data to predict demographic attributes and study homophily with respect to those attributes. Furthermore, data are too revealing in terms of privacy but, at the same time, do not provide enough information for researchers~\cite{Golder2014}. Individuals can be identified in anonymised social networks~\cite{Backstrom2011, Narayanan2008}, and augmenting the network data with demographic information would make re-identification even easier.

@@

% In this paper, we aim to address three challenges. First, numerous studies~\cite{Cheadle2012, Currarini2009, Hipp2009,Marsden1987, Marsden1988, McPherson2001, Smith2014, Wimmer2010} refer to Blau's~\cite{Blau1977} intuitive and powerful theory of social structure: people occupy a multidimensional social space spanned by their demographic attributes, and their rates of interaction depend on their relative positions. However, to the best of our knowledge, only \textcite{Hipp2009} explicitly consider such a social space---also referred to as ``Blau space''~\cite{McPherson2001}.

@@

We derive the posterior for parameters of the conditionally-independent edge model given partial observations of social networks obtained from surveys% and show that the posterior is equivalent to the widely-used ``prior correction'' when the probability for two people to connect is modelled by a logistic model
. We apply our approach to nine existing datasets from the United Kingdom (UK) and two from the United States (US). Our analysis reveals that the effects of homophily on society are remarkably stable in both countries regardless of time and the specific nature of relationships. Using the suite of segregation measures, we find that physical separation and age are the most important factors contributing to the segregation of society. We provide recommendations for conducting surveys to infer homophily in social networks and discuss future work, including principled imputation of demographic attributes in social networks and using the generative model to generate synthetic networks: an approach that has attracted interest to circumvent the difficulties associated with anonymising social network data~\cite{Pfeiffer-III2014, Lieberman2010, Lieberman2012, Nettleton2016, Sathanur2017}.

@@

\subsection{Generative network model\label{sec:model}}

% \begin{figure}
%     \includegraphics{survey-synthetic}
%     \titlecaption{A simple generative model for ego networks obtained from surveys.}{Panel~(a) shows synthetic survey data generated by our model. Seeds $\seeds$ are represented by blue squares, nominees $\nominees$ by orange circles, and the remainder of the population $\missing$ by grey circles. Panel~(b) illustrates the parts of the adjacency matrix revealed by a particular two-stage survey. Connections are represented by black entries in the adjacency matrix, absent connections by white entries, and lacking information by grey entries.\label{fig:survey-synthetic}}
% \end{figure}

We consider a generative model for social networks for a population of $n$ individuals $\population$ who occupy a Blau space $\blauspace$ spanned by their demographic attributes, such as age, income, or gender. In contrast to common latent space models~\cite{Hoff2002, Hoff2008, Handcock2007}, the attributes are observed, although \textcite{Hoff2002} also consider an extension including covariates. The $q$-dimensional attribute vector $x_i\in\blauspace$ for each individual $i\in \population$ is drawn independently from a distribution of demographic attributes $P(x_i)$. Connections between individuals are encoded by the binary adjacency matrix $A$ such that $A_{ij}=1$ if $j$ considers $i$ to be a friend and $A_{ij}=0$ otherwise. We assume that people do not interact with themselves such that $A_{ii}=0$ for all $i$, and that connections are directed because social ties need not be reciprocated. For example, Alice may consider Bob to be a friend whereas the reverse need not be true~\cite{Ball2013}.

Given the positions of two individuals $i$ and $j$ in Blau space, we assume that connections form independently with probability $\rho(x_i, x_j)$, i.e.\ edges are conditionally-independent given the attributes of nodes~\cite{Fienberg2012}. The assumption of conditionally-independent edges can be problematic. For example, it is difficult to reproduce some of the statistics of real-world networks, such as heavy-tailed degree distributions, if the node density is homogeneous~\cite{Barnett2007}. Furthermore, the average degree scales linearly with the number of nodes unless the connectivity kernel $\rho$ is adjusted to compensate~\cite{Caron2017}. Nevertheless, we use spatial network models because the connectivity kernel is easily interpretable, and they are able to capture salient features of social networks. For example, nodes in high-density regions have larger degrees on average~\cite{Barnett2007}. Similarly, members of the ethnic majority have more social ties in social networks in US high schools~\cite{Currarini2009}.

% We now turn to the data collection process. Observations are collected in two stages: first, we sample a subset $\seeds\subset\population$ of \emph{seeds} uniformly at random and collect information about their demographic background. We also ask each seed $j\in \seeds$ about their social ties $\nominees_j=\{i\in\population:A_{ij}=1\}$, henceforth \emph{nominees}. Second, we collect demographic information about the nominees $\nominees=\bigcup_{j\in \seeds} \nominees_j$ by asking the seeds about their friends~\cite{Huckfeldt1983, McPherson2006, Kalmijn2007} or conducting follow-up surveys~\cite{Johnson1989}.

% Panel~(a) of \cref{fig:survey-synthetic} illustrates the model with a simple example: we first draw the positions $x$ of $n=1,000$ nodes in the social space uniformly at random from the two-dimensional unit box. Second, we connect nodes with each another using a connectivity kernel that induces homophily along the two dimensions of Blau space. Finally, we sample $\card{\seeds}=100$ seeds from the population, and collect information about their demographic background as well as their $\card{\nominees}=108$ nominees, which allows us to embed the ego networks in Blau space.

\subsection{Developing a model-based segregation measure}

In addition to addressing the challenges mentioned in \cref{sec:introduction}, a social segregation measure should satisfy the following properties: first, the measure should be insensitive to the overall edge density to facilitate comparison of segregation across different networks. Otherwise, the segregation measure would depend on the size of the population because the edge density scales as $n^{-1}$ if the average degree is approximately constant. Second, following \textcite{Freeman1978}, we would like the measure to capture the notion that segregation places ``restrictions on the access of people to one another''. Third, the measure should be easily interpretable, and it should have a natural notion of the absence of segregation when individuals form connections without regard to their positions in Blau space. For example, the difference of within- and between-group ties considered by \textcite{Krackhardt1988} depends on the sizes of the groups even if there is no homophily: there is no natural reference point.

A single measure cannot capture the complexities of social networks, and we will develop a suite of measures applicable at different scales: the segregation or separation between any two individuals, the isolation experienced by any one individual, and a global measure of segregation. Starting at the microscopic level, we use the generative model to define a \emph{pairwise} measure of separation between two individuals with attributes $x$ and $y$:
\begin{align}
    \varphi(x, y) &= \logit \rho(y,y) - \logit\rho(x,y),\label{eq:segregation-pairwise}\\
    \text{where }\logit\rho &= \log\left(\frac{\rho}{1-\rho}\right)\nonumber
\end{align}
are the log odds for a connection to form with probability $\rho$. The probability $\rho(y, y)$ for an individual with attributes $y$ to connect with someone with identical attributes serves as a reference point: the measure does not depend on the overall edge density. The pairwise measure $\varphi$ may be understood as the social isolation experienced by $x$ as a result of the behaviour of $y$. The measure is zero if people do not discriminate with respect to the attributes or if two individuals have the same demographic attributes.
% In a heterophilous society, the measure is negative for any pair of individuals.
For a homophilous connectivity kernel, the measure is positive and is a \emph{semi-metric} for Blau space; we will consider a family of connectivity kernels for which $\varphi$ is a true metric in \cref{sec:segregation-distance}.

\begin{prop}\label{prop:segregations-semi-metric}
    If the connectivity kernel is homophilous, symmetric, and homogeneous, the pairwise segregation measure is a \emph{semi-metric}~\cite{Wilson1931}: it satisfies the properties of a metric, including non-negativity, the identity of indiscernibles, and symmetry with respect to exchange of arguments---except the triangle inequality.
\end{prop}
\begin{proof}
    First, the measure is non-negative because homophily implies that $\rho(x, y)<\rho(y, y)$ and $\logit$ is a monotonic function. Second, the measure is zero for any two individuals with the same attributes by substitution into \cref{eq:segregation-pairwise}. Third, the first term of \cref{eq:segregation-pairwise} is constant, and the second is invariant under exchange of the arguments because the kernel is homogeneous and symmetric, respectively.
\end{proof}

The less likely two people are to connect, the larger the social distance between them. The assumptions required for \cref{prop:segregations-semi-metric} to hold may seem restrictive, but they are satisfied by most studies of spatial networks~\cite{Barnett2007,Lambiotte2008,Butts2012,Expert2011}.

@@

For example, consider a stochastic block model (SBM)~\cite{Snijders2011} with intra-group connection probability $\rho_\mathrm{same}$ and inter-group connection probability $\rho_\mathrm{different}<\rho_\mathrm{same}$. The pairwise segregation measure
\[
    \varphi(x, y) = \left(1-\delta_{xy}\right)\left(\logit\rho_\mathrm{same}-\logit\rho_\mathrm{different}\right),
\]
where $x$ and $y$ denote the blocks and $\delta_{xy}$ is the Kronecker delta. The segregation measure only depends on block membership, and it is not affected by the size of each block. The pairwise measure for members of different blocks is the difference of log odds ratios for the existence of intra-group ties as opposed to inter-group ties previously considered by \textcite{Moody2001}.

% Both the connectivity kernel and the pairwise measure convey information about how two individuals connect with one another.
To quantify segregation at the level of an individual $x$, we define the \emph{individual} segregation measure or \emph{social isolation}
\begin{equation}
    \phi(x) = \int dy \ P(y) \varphi(x, y),\label{eq:segregation-individual}
\end{equation}
which measures the average distance from members of society to $x$. For the SBM,
\begin{equation}
    \phi(x) = (1-P(x))\left(\logit\rho_\mathrm{same}-\logit\rho_\mathrm{different}\right),\label{eq:segregation-individual-sbm}
\end{equation}
and members of all blocks experience the same degree of isolation if the blocks are of the same size. If the sizes are unequal, minorities experience more isolation and majority groups experience less isolation. Indeed, ethnic minorities in schools tend to be more isolated and have fewer social ties~\cite{Currarini2009}.

@@

Some of the demographic attributes of seeds and nominees are missing because respondents did not know or refused to provide the information, and we drop individuals with one or more missing attributes, as shown in \cref{tbl:survey-sample-size}. % Such a complete-case analysis can introduce biases if the data are not missing completely at random~\cite{Rubin1976}, but handling the missing data in a principled fashion would, once again, require us to develop a model for demographic attributes~\cite{Pigott2001, Gelman2013}.
